---
title: "One-line Monte Carlo simulation"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_download: true
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# License

[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/deed.en)

```{r, include=FALSE}

# set default chunk options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# One line Monte Carlo simulation

Answering the question "what is the distribution of *p*-values under the null hypothesis?"

```{r}

set.seed(42)

hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value))

```

This is a complete Monte Carlo simulation. But how is it constructed? Let's built it up more slowly.

# Generate normally distributed data

Draw data from a normally distributed population and plot it using `rnorm()`

```{r}

set.seed(42)

rnorm(n = 50, m = 0, sd = 1) |> 
  hist()

```

# Generate two normally distributed data sets and fit a *t*-test 

Note the population means are equivalent. What does a *t*-test test for?

```{r}

set.seed(42)

t.test(rnorm(n = 50, m = 0, sd = 1), 
       rnorm(n = 50, m = 0, sd = 1))

```

# Generate two normally distributed data sets and fit a *t*-test and extract the *p*-value

```{r}

set.seed(42)

t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value

```

# Do this generate-analyze-extract many times

Using the `replicate()` function, and plot the *p*-values using `hist()`.

```{r}

set.seed(42)

hist(replicate(1000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value))

```

# Increase the number of iterations 

Why is the distribution more uniform?

```{r}

set.seed(42)

hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value))

```

# Compare with distribution of *p*-values when population means are not equal

'Small', 'medium' and 'large' differences in population means.

```{r}

set.seed(42)

hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0.2, sd = 1))$p.value))
hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0.5, sd = 1))$p.value))
hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0.8, sd = 1))$p.value))

```

# Check your learning

- What are the five core components of a Monte Carlo simulation?
- Can you mentally match these five components onto the above code (sometimes in different chunks!) to understand how, together, they answer the question "what is the distribution of *p*-values under the null vs. alternative hypothesis?"

# Session info

```{r}

sessionInfo()

```





