---
title: "Learning simulation studies"
subtitle: "Lesson 2"
author: "Template by Ian Hussey, additions by [student's name]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen=999)

```

```{r dependencies and settings}

# dependencies
library(tidyverse)
library(purrr)
library(scales)
library(knitr)
library(kableExtra)
library(janitor)
# install.packages("devtools")
# library(devtools)
# devtools::install_github("ianhussey/simulateR")
library(simulateR) # available from github - uncomment the above lines to install

# set seed
set.seed(42)

```

# Parameter recovery of $\mu$ and $\sigma$

The previous lesson finished with code for a simulation to recover $\mu$ and $\sigma$ across a large number of iterations of smaller samples. Here is the code again:

```{r}

set.seed(42)

# new values 
n_samples <- 100
n_iterations <- 10000
mu <- 0.33
sigma <- 1

# create two new results vectors
results_means <- numeric(n_iterations)
results_sds <- numeric(n_iterations)

for(i in 1:n_iterations){
  # generate data sampled from a normal population using rnorm
  simulated_scores <- 
    rnorm(n = n_samples, 
          mean = mu, 
          sd = sigma)
  
  # compute the mean for this simulation and store it 
  # in the `i`th element of each results vector
  results_means[i] <- mean(simulated_scores)
  results_sds[i] <- sd(simulated_scores)
}

# compute the mean of means
mean(results_means) |> round_half_up(2)
# check that the mean of sample means is equal to the population mean (mu)
mean(results_means) |> round_half_up(2) == mu

# compute the mean of SDs
mean(results_sds) |> round_half_up(2)
# check that the mean of sample SDs is equal to the population SD (sigma)
mean(results_sds) |> round_half_up(2) == sigma

```

We also noted that it is important to vary the simulation parameters. **In the context of a simulation, this is what constitutes an experiment**: parameters are systematically varied, and results are observed. Publications reporting such results are typically called "simulation studies", but it might be more accurate to call (at least some of them) "simulation experiments". 

Varying simulation parameters in order to observe results can (a) provide evidence that (elements of) your simulation are working correctly, and/or (b) provide insights into the thing you are varying and how it changes across input parameters or other conditions.

# Running simulations with different parameters more easily

**sth about instructing the user to run the above multiple times with different inputs, or showing long-hand that examining different parameters values requires you to duplicate the code a lot.**

In lesson 1, we used a for loop to run be able to easily run an arbitrary number of iterations of a simulation. 

Comparably, if we want to run the same simulation an arbitrary number of times, perhaps using different population parameters, maybe there are other changes we could make to the code to make it possible to do this.

To do this, we'll need to first take a step back and talk about functions.

# Functions

Writing R code involves making frequent use of functions. Functions take inputs, do things with them or based on them, and return outputs. 

`mean()` is a variable. It takes a vector of numeric values (its input) and returns their average. 

`for()` is a variable,

# Writing functions

We can write our own functions too, when there isn't one that does exactly what we want. Functions take "arguments", the variables specified inside the round brackets `()`, and use these in the code inside the curly brackets `{}`, and return whatever is specified by `return()`.

Note that once this code is run, the function appears in your environment. Your environment isn't just for data!

```{r}

multiply_by_two <- function(x){
  res <- x*2 # note that "res" is common but not required; its short for result
  return(res)
}

multiply_by_two(1)
multiply_by_two(20)

```

`return()` does not have to be specified explicitly (in replicate or other functions): if the last line would print something, it returns it. However, implicit behavior is usually more confusing, so try to avoid relying on it.

```{r}

multiply_by_two <- function(x){
  x*2
}

multiply_by_two(1)
multiply_by_two(20)

```

Note that the syntax for functions can differ. I wouldn't advise specfiying it this way without curly brackets, but but know that it can be done.

```{r}

multiply_by_two <- function(x)x*2

multiply_by_two(1)
multiply_by_two(20)

```

Recently, other syntax for functions was added to R. Again, I wouldn't advise specfiying it this way, but know that it can be done.

```{r}

multiply_by_two <- \(x)x*2

multiply_by_two(1)
multiply_by_two(20)

```

# Convert our simulation using for loops into a function

Here's the original code to do parameter recovery for the mean of a normal distribution. 

```{r}

set.seed(42)

# new values 
n_samples <- 100
n_iterations <- 10000
mu <- 0.33
sigma <- 1

# create two new results vectors
results_means <- numeric(n_iterations)
results_sds <- numeric(n_iterations)

for(i in 1:n_iterations){
  # generate data sampled from a normal population using rnorm
  simulated_scores <- 
    rnorm(n = n_samples, 
          mean = mu, 
          sd = sigma)
  
  # compute the mean for this simulation and store it 
  # in the `i`th element of each results vector
  results_means[i] <- mean(simulated_scores)
}

# compute the mean of means
mean(results_means) |> round_half_up(2)
# check that the mean of sample means is equal to the population mean (mu)
mean(results_means) |> round_half_up(2) == mu

```

Let's convert the simulation code into a very simple function which, for now, has no arguments. 

```{r}

set.seed(42)

sim <- function(){
  # create two new results vectors
  res <- numeric(10000)
  
  for(i in 1:10000){
    # generate data sampled from a normal population using rnorm
    dat <- rnorm(n = 100, 
                 mean = 0.33, 
                 sd = 1.00)
    
    # compute the mean for this simulation and store it 
    # in the `i`th element of each results vector
    res[i] <- mean(dat)
  }
  return(res)
}

sim() |> 
  mean() |> 
  round_half_up(2)

```

We can make the function more abstract by using arguments

```{r}

set.seed(42)

sim <- function(n_iterations, n_samples, mu, sigma){
  # create two new results vectors
  res <- numeric(n_iterations)
  
  for(i in 1:n_iterations){
    # generate data sampled from a normal population using rnorm
    dat <- rnorm(n = n_samples, 
                 mean = mu, 
                 sd = sigma)
    
    # compute the mean for this simulation and store it 
    # in the `i`th element of each results vector
    res[i] <- mean(dat)
  }
  return(res)
}

sim(n_iterations = 10000,
    n_samples = 100,
    mu = 0.33,
    sigma = 1.00) |> 
  mean() |> 
  round_half_up(2)

sim(n_iterations = 10000,
    n_samples = 100,
    mu = 0.76,
    sigma = 1.00) |> 
  mean() |> 
  round_half_up(2)

```

# Abstraction

If I was sitting down to write a new simulation from scratch, I would not start by writing a function like in the above chunk. I would start with getting rnorm() to work, then puttting it inside a for loop, and then converting it to a function. When you read someone else's  

**make very clear that this process of abstracting and iterating is how you write this stuff. you dont start writing "function()" and just know what youre doing.**

# Alternatives to for loops

For loops are no the only way to implement iterations.

## (s)apply()

Instead of for loops, you can also use `sapply()`, which is one of multiple `apply()` functions in R. 

```{r}

set.seed(42)

res_sim <- sapply(1:10000, function(i) {
  dat <- rnorm(n = 100, 
               mean = 0.33, 
               sd = 1.00)
  res <- mean(dat)
  return(res)
})

res_sim |> 
  mean() |> 
  round_half_up(2)

```

## replicate()

Instead of for loops, you can also use `replicate()`. 

```{r}

set.seed(42)

res_sim <- replicate(10000, {
  dat <- rnorm(n = 100, 
               mean = 0.33, 
               sd = 1.00)
  res <- mean(dat)
  return(res)
})

res_sim |> 
  mean() |> 
  round_half_up(2)

```

# Convert `replicate()` code to function

So that it can be run an aribtrary number of times.

First, lets replace the hard-coded parameters with variables.

```{r}

set.seed(42)

n_samples <- 100
n_iterations <- 10000
mu <- 0.33
sigma <- 1

sim1 <- replicate(n_iterations, {
  dat <- rnorm(n = n_samples, 
               mean = mu, 
               sd = sigma)
  res <- mean(dat)
  return(res)
})

sim1 |> mean() |> round_half_up(2)

```

Rewrite as a function. Use default arguments for everything but n_samples.

```{r}

set.seed(42)

simulation_1 <- function(n_samples, n_iterations = 10000, mu = 0.33, sigma = 1.00){
  
  iteration_res <- replicate(n_iterations, {
    dat <- rnorm(n = n_samples, 
                 mean = mu, 
                 sd = sigma)
    res <- mean(dat)
    return(res)
  })
  
  iteration_res |> mean()
}

simulation_1(n_samples = 10)
simulation_1(n_samples = 100)

```

What if we want to abstract even more? E.g., to apply the function to a range of values of n_samples in sequence.

The {purrr} library has a number of different `map` functions that allow you to do this, i.e., to "map" a function onto a range of input values.

```{r}

set.seed(42)

sample_sizes <- c(5, 10, 25, 50, 100, 1000)

map_dbl(sample_sizes, simulation_1) |>
  round_half_up(digits = 4)

```

Maybe a vector is not the most useful way to store the output. Maybe a data frame would be more useful, in order to analyze and plot results. We can also map within a mutate call:

```{r}

set.seed(42)

sample_sizes <- c(5, 10, 25, 50, 100, 1000)

dat <- 
  data.frame(sample_size = sample_sizes) |>
  mutate(mean_mean = map_dbl(sample_size, simulation_1))

dat

```

We could greatly increase the number of conditions studied by just changing the samples_sizes vector:

```{r}

set.seed(42)

sample_sizes <- seq(from = 10, to = 500, by = 10)

sample_sizes

dat <- 
  data.frame(sample_size = sample_sizes) |>
  mutate(mean_mean = map_dbl(sample_size, simulation_1))

ggplot(dat, aes(sample_size, mean_mean)) +
  geom_hline(yintercept = 0.33, linetype = "dotted") +
  geom_point() +
  theme_linedraw()

```

# Check your understanding

- Think about what is actually being done in the last simulation. In lesson 1, we started by generating data from a normal distribution using rnorm, and calculate its mean. We had to re run the chunk each time we wanted to generate a new dataset and mean. We wrote a for loop (or apply, or replicate) to do this part an arbitrary number of times with a single click. We wrapped this in a function to be able to change the inputs and run the same code an arbitrary number of times without repeating the loop code. We then used map to map the function onto a range of inputs in a single line of code. How many times would you have had to manually run the original rnorm code to produce the output in the final plot, if you were doing this all manually?
- This simulation is still limited, as it only produces the results (the mean of means for that sample size). It can't return multiple estimates (eg the mean and also SD), and it only returns the summary (the mean of means) and not a) each simulated dataset or b) the mean of each simulated dataset.  

# Session info

```{r}

sessionInfo()

```




