---
title: "The impact of careless responding on correlations"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---


```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

# Overview

Simulations to try to conceptually reproduce the effects described by [Stosic et al. 2024](https://journals.sagepub.com/doi/epub/10.1177/25152459241231581) and their [shiny app](https://fuhred.shinyapps.io/CarelessRespondingSimulator/).

# TODO 

- outline assumptions, ie careless responding = uniform distribution; bounded response options + censoring.
- investigate how the simulations handle generating data for multiple items, and how it converts to likert or bounded data. The simulation requires that the careless responding has a uniform distribution and therefore bounds, so this needs handling in one way or another. i could maybe use faux::norm2likert but i don't understand it well enough. faux::norm2trunc could also be useful - could ignore the interval nature of the responses but still capture their bounding.
- change each sim and their plots to the ones now in truncated normal 

```{r fig.height=5, fig.width=10}

# remove all objects from environment ----
#rm(list = ls())


# dependencies ----
# repeated here for the sake of completeness 

library(tidyr)
library(dplyr)
library(tibble)
library(forcats)
library(readr)
library(purrr) 
library(ggplot2)
library(janitor)
library(knitr)
library(kableExtra)
library(faux)
library(lavaan)

# set the seed ----
# for the pseudo random number generator to make results reproducible
set.seed(42)

# define data generating function ----
generate_data <- function(n,
                          prob_careless,
                          rho_careful,
                          mu_x_careful,
                          mu_y_careful) { 

  n_careless <- floor(n * prob_careless)
  n_careful <- n - n_careless
  
  data_careful <- 
    faux::rnorm_multi(n = n_careful, 
                      mu = c(y = mu_y_careful, x = mu_x_careful), 
                      sd = c(1, 1), 
                      r = matrix(c(1, rho_careful, 
                                   rho_careful, 1), 
                                 ncol = 2)) |>
    mutate(type = "careful") |>
    # convert to likert
    # mutate(y = as.numeric(as.character(cut(y, breaks = 7, labels = 1:7))),
    #        x = as.numeric(as.character(cut(x, breaks = 7, labels = 1:7))))
    mutate(y = janitor::round_half_up(y, 2),
           x = janitor::round_half_up(x, 2)) |>
    mutate(y = case_when(y < -3 ~ -3, 
                         y > +3 ~ +3,
                         TRUE ~ y),
           x = case_when(x < -3 ~ -3, 
                         x > +3 ~ +3,
                         TRUE ~ x))
    # mutate(y = as.integer(y + 4),
    #        x = as.integer(x + 4))

  data_careless <-
    # data.frame(y = sample(-3:+3, size = n_careless, replace = TRUE),
    #            x = sample(-3:+3, size = n_careless, replace = TRUE)) |>
    data.frame(y = janitor::round_half_up(runif(n = n_careless, min = -3, max = +3), digits = 2),
               x = janitor::round_half_up(runif(n = n_careless, min = -3, max = +3), digits = 2)) |>
    mutate(type = "careless") 
  
  data <- bind_rows(data_careful,
                    data_careless) |>
    rownames_to_column(var = "id")
  
  return(data)
}


# define data analysis function ----
analyse_data <- function(data) {
  
  fit_all <- cor.test(data$y,
                      data$x, 
                      method = "pearson")
  
  dat_careful <- data |>
    filter(type == "careful")
  
  fit_careful <- cor.test(dat_careful$y,
                          dat_careful$x, 
                          method = "pearson")
  
  dat_careless <- data |>
    filter(type == "careless")
  
  fit_careless <- cor.test(dat_careless$y,
                           dat_careless$x, 
                           method = "pearson")
  
  results <- tibble(r_all = fit_all$estimate,
                    p_all = fit_all$p.value,
                    r_careful = fit_careful$estimate,
                    p_careful = fit_careful$p.value,
                    r_careless = fit_careless$estimate,
                    p_careless = fit_careless$p.value,)
  
  return(results)
}

```

# Just noise (negative control, tracks intuition)

Both careful and careless responding have means at midpoint of scale, population correlation is non-zero. 

\TODO **Results shouldn't be biased from the population value, but seem to be? Recheck the paper to see if I understand what should happen when its just noise and no difference in means.**

```{r fig.height=6, fig.width=6}

# define experiment parameters ----
experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = 0.15,
  rho_careful = 0.5, 
  mu_x_careful = 0,
  mu_y_careful = 0,
  iteration = 1:1000
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))
  
# check data using iteration 400 (randomly chosen)
dat <- simulation$generated_data[[400]] |>
  pivot_longer(cols = c(x, y))

ggplot(dat, aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(name ~ type, scales = "free") +
  theme_linedraw() +
  coord_cartesian(xlim = c(-3.5, +3.5))

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point(aes(color = type, shape = type)) +
  geom_smooth(aes(color = type), method = "lm", alpha = 0) +
  #geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15)) +
  ggtitle("By responding type")

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  theme_linedraw() +
  ggtitle("Effect without excluding careless responders")

# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n,
           prob_careless,
           rho_careful,
           mu_x_careful,
           mu_y_careful) |>
  summarize(mean_r = mean(r_all),
            mean_r_careful = mean(r_careful),
            mean_r_careless = mean(r_careless),
            bias_mean_diff = mean(r_all - rho_careful),
            #bias_mean_proportion = mean(bias_mean_diff/r),
            .groups = "drop")

# print table
simulation_summary |>
  #arrange(bias_mean_proportion) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Create effect

Mean in the careful responding diverges from the midpoint of the scale (in same direction on X and Y variables), population correlation is 0.

## Mean below the middle point

```{r fig.height=6, fig.width=6}

# define experiment parameters ----
experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = 0.15,
  rho_careful = 0, 
  mu_x_careful = -2,
  mu_y_careful = -2,
  iteration = 1:1000
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))
  
# check data using iteration 400 (randomly chosen)
dat <- simulation$generated_data[[400]] |>
  pivot_longer(cols = c(x, y))

ggplot(dat, aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(name ~ type, scales = "free") +
  theme_linedraw() +
  coord_cartesian(xlim = c(-3.5, +3.5))

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point(aes(color = type, shape = type)) +
  geom_smooth(aes(color = type), method = "lm", alpha = 0) +
  #geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15)) +
  ggtitle("By responding type")

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  theme_linedraw() +
  ggtitle("Effect without excluding careless responders")

# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n,
           prob_careless,
           rho_careful,
           mu_x_careful,
           mu_y_careful) |>
  summarize(mean_r = mean(r_all),
            mean_r_careful = mean(r_careful),
            mean_r_careless = mean(r_careless),
            bias_mean_diff = mean(r_all - rho_careful),
            #bias_mean_proportion = mean(bias_mean_diff/r),
            .groups = "drop")

# print table
simulation_summary |>
  #arrange(bias_mean_proportion) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

## Mean above the middle point

```{r fig.height=6, fig.width=6}

# define experiment parameters ----
experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = 0.15,
  rho_careful = 0, 
  mu_x_careful = +2,
  mu_y_careful = +2,
  iteration = 1:1000
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))
  
# check data using iteration 400 (randomly chosen)
dat <- simulation$generated_data[[400]] |>
  pivot_longer(cols = c(x, y))

ggplot(dat, aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(name ~ type, scales = "free") +
  theme_linedraw() +
  coord_cartesian(xlim = c(-3.5, +3.5))

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point(aes(color = type, shape = type)) +
  geom_smooth(aes(color = type), method = "lm", alpha = 0) +
  #geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15)) +
  ggtitle("By responding type")

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  theme_linedraw() +
  ggtitle("Effect without excluding careless responders")

# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n,
           prob_careless,
           rho_careful,
           mu_x_careful,
           mu_y_careful) |>
  summarize(mean_r = mean(r_all),
            mean_r_careful = mean(r_careful),
            mean_r_careless = mean(r_careless),
            bias_mean_diff = mean(r_all - rho_careful),
            #bias_mean_proportion = mean(bias_mean_diff/r),
            .groups = "drop")

# print table
simulation_summary |>
  #arrange(bias_mean_proportion) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Inflate effect

Mean in the careful responding diverges from the midpoint of the scale (in same direction on X and Y variables), population correlation is non-zero.

```{r fig.height=6, fig.width=6}

# define experiment parameters ----
experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = 0.15,
  rho_careful = 0.5, 
  mu_x_careful = 2,
  mu_y_careful = 2,
  iteration = 1:1000
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))
  
# check data using iteration 400 (randomly chosen)
dat <- simulation$generated_data[[400]] |>
  pivot_longer(cols = c(x, y))

ggplot(dat, aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(name ~ type, scales = "free") +
  theme_linedraw() +
  coord_cartesian(xlim = c(-3.5, +3.5))

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point(aes(color = type, shape = type)) +
  geom_smooth(aes(color = type), method = "lm", alpha = 0) +
  #geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15)) +
  ggtitle("By responding type")

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  theme_linedraw() +
  ggtitle("Effect without excluding careless responders")

# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n,
           prob_careless,
           rho_careful,
           mu_x_careful,
           mu_y_careful) |>
  summarize(mean_r = mean(r_all),
            mean_r_careful = mean(r_careful),
            mean_r_careless = mean(r_careless),
            bias_mean_diff = mean(r_all - rho_careful),
            #bias_mean_proportion = mean(bias_mean_diff/r),
            .groups = "drop")

# print table
simulation_summary |>
  #arrange(bias_mean_proportion) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Dilute effect

Mean in the careful responding diverges from the midpoint of the scale (in opposite directions on X and Y variables), population correlation is non-zero.

```{r fig.height=6, fig.width=6}

# define experiment parameters ----
experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = 0.15,
  rho_careful = 0.5, 
  mu_x_careful = +2,
  mu_y_careful = -2,
  iteration = 1:1000
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))
  
# check data using iteration 400 (randomly chosen)
dat <- simulation$generated_data[[400]] |>
  pivot_longer(cols = c(x, y))

ggplot(dat, aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(name ~ type, scales = "free") +
  theme_linedraw() +
  coord_cartesian(xlim = c(-3.5, +3.5))

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point(aes(color = type, shape = type)) +
  geom_smooth(aes(color = type), method = "lm", alpha = 0) +
  #geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15)) +
  ggtitle("By responding type")

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(-3.5, +3.5)) +
  theme_linedraw() +
  ggtitle("Effect without excluding careless responders")

# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n,
           prob_careless,
           rho_careful,
           mu_x_careful,
           mu_y_careful) |>
  summarize(mean_r = mean(r_all),
            mean_r_careful = mean(r_careful),
            mean_r_careless = mean(r_careless),
            bias_mean_diff = mean(r_all - rho_careful),
            #bias_mean_proportion = mean(bias_mean_diff/r),
            .groups = "drop")

# print table
simulation_summary |>
  #arrange(bias_mean_proportion) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Dev

## item level data with measurement error

solution using lavaan to generate item level data with measurement error that is reduced to a participant level mean score.

```{r}

generate_data <- function(rho,
                          n,
                          item_loading_x = .8, 
                          item_loading_y = .8, 
                          n_items_x = 10,
                          n_items_y = 10){
  
  if(rho > 1 |
     rho < 0) stop("rho must be > 0 and < 1")
  
  if(item_loading_x > .99 |
     item_loading_x < .01 |
     item_loading_y > .99 | 
     item_loading_y < .01) stop("item loadings must be >= 0.01 and <= 0.99")
  
  generate_lavaan_syntax_latent <- function(variable_name, item_loading, n_items){
    paste(variable_name, 
          " =~ " , 
          paste(item_loading, 
                "*", 
                variable_name, 
                "_", 
                seq(1, n_items), 
                sep = "", 
                collapse = " + "), 
          sep = "")
  }
  
  population_model <- 
    paste(
      paste0("y_latent ~ ", rho, "*x_latent"),
      generate_lavaan_syntax_latent(variable_name = "y",
                                    item_loading = item_loading_y,
                                    n_items = n_items_y),
      generate_lavaan_syntax_latent(variable_name = "x",
                                    item_loading = item_loading_x,
                                    n_items = n_items_x),
      sep = "; \n "
    )
  
  data <- 
    lavaan::simulateData(model = population_model, 
                         sample.nobs = n) |>
    # calculate mean scores for each participant
    rownames_to_column(var = "id") |>
    ## make longer
    pivot_longer(cols = -id,
                 names_to = c("scale", "item"),
                 names_pattern = "(.)_(.)",
                 values_to = "score") |> # calculate means
    group_by(id, scale) |>
    dplyr::summarize(mean_score = mean(score), .groups = "keep") |>
    ungroup() |>
    ## make wider again
    pivot_wider(id_cols = id,
                names_from = scale,
                #names_glue = "{scale}_meanscore",
                names_glue = "{scale}",
                values_from = mean_score)
  
  return(data)
}

generate_data(rho = 0.5, n = 800)


```

## truncated normal data

```{r}

x <- rnorm(1000000)
y <- norm2trunc(x, 
                min = 1, 
                max = 7, 
                mu = 4, 
                sd = 1)
mean(y)
sd(y)
g <- ggplot2::ggplot() + ggplot2::geom_point(ggplot2::aes(x, y))
ggExtra::ggMarginal(g, type = "histogram")


x <- rnorm(1000000)
y <- norm2trunc(x, min = 1, max = 7, mu = 5.5, sd = 1)
mean(y)
sd(y)
g <- ggplot2::ggplot() + ggplot2::geom_point(ggplot2::aes(x, y))
ggExtra::ggMarginal(g, type = "histogram")

# y <- norm2likert(x, 
#                  prob = c(.1, .2, .4, .8, .4, .2, .1), 
#                  mu = 4, 
#                  sd = 1)
# g <- ggplot2::ggplot() + ggplot2::geom_point(ggplot2::aes(x, y))
# ggExtra::ggMarginal(g, type = "histogram")

```

```{r fig.height=6, fig.width=6}

generate_data <- function(n,
                          prob_careless,
                          rho_careful,
                          mu_x_careful,
                          mu_y_careful) { 

  n_careless <- floor(n * prob_careless)
  n_careful <- n - n_careless
  
  data_careful <- 
    faux::rnorm_multi(n = n_careful, 
                      mu = c(y = 1, x = 1), 
                      sd = c(1, 1), 
                      r = matrix(c(1, rho_careful, 
                                   rho_careful, 1), 
                                 ncol = 2)) |>
    mutate(type = "careful") |>
    # convert to likert
    mutate(x = faux::norm2trunc(x, min = 1, max = 7, mu = mu_x_careful, sd = 1),
           y = faux::norm2trunc(y, min = 1, max = 7, mu = mu_y_careful, sd = 1))

  data_careless <-
    data.frame(x = runif(n = n_careless, min = 1, max = 7),
               y = runif(n = n_careless, min = 1, max = 7)) |>
    mutate(type = "careless") 
  
  data <- bind_rows(data_careful,
                    data_careless) |>
    rownames_to_column(var = "id")
  
  return(data)
}

# define data analysis function ----
analyse_data <- function(data) {
  
  fit_all <- cor.test(data$y,
                      data$x, 
                      method = "pearson")
  
  dat_careful <- data |>
    filter(type == "careful")
  
  fit_careful <- cor.test(dat_careful$y,
                          dat_careful$x, 
                          method = "pearson")
  
  dat_careless <- data |>
    filter(type == "careless")
  
  fit_careless <- cor.test(dat_careless$y,
                           dat_careless$x, 
                           method = "pearson")
  
  results <- tibble(r_all = fit_all$estimate,
                    p_all = fit_all$p.value,
                    r_careful = fit_careful$estimate,
                    p_careful = fit_careful$p.value,
                    r_careless = fit_careless$estimate,
                    p_careless = fit_careless$p.value,)
  
  return(results)
}

# set seed
set.seed(42)

# define experiment parameters ----
prob_careless <- 0.15

experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = prob_careless,
  rho_careful = 0, 
  mu_x_careful = 2,
  mu_y_careful = 2,
  iteration = 1:1000
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))
  


# check data using iteration 400 (randomly chosen)
dat <- simulation$generated_data[[400]] |>
  pivot_longer(cols = c(x, y))

ggplot(dat, aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(name ~ type, scales = "free") +
  theme_linedraw() +
  coord_cartesian(xlim = c(1, 7))

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point(aes(color = type, shape = type)) +
  geom_smooth(aes(color = type), method = "lm", alpha = 0) +
  #geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(1, 7)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15)) +
  ggtitle("By responding type")

ggplot(simulation$generated_data[[400]], aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", alpha = 0, color = "black") +
  coord_cartesian(xlim = c(1, 7)) +
  theme_linedraw() +
  ggtitle("Effect without excluding careless responders")

# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n,
           prob_careless,
           rho_careful,
           mu_x_careful,
           mu_y_careful) |>
  summarize(mean_r = mean(r_all),
            mean_r_careful = mean(r_careful),
            mean_r_careless = mean(r_careless),
            bias_mean_diff = mean(r_all - rho_careful),
            #bias_mean_proportion = mean(bias_mean_diff/r),
            .groups = "drop")

# print table
simulation_summary |>
  #arrange(bias_mean_proportion) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```


```{r fig.height=5}

plucked_iteration <- 400

dat_all <- 
  bind_rows(simulation$generated_data[[plucked_iteration]] |>
              mutate(type = case_when(type == "careful" ~ "True effect\n(careful responding)",
                                      type == "careless" ~ "Noise\n(careless responding)")),
            simulation$generated_data[[plucked_iteration]] |>
              mutate(type = "Observed effect\n(careful + careless)")) |>
  mutate(type = fct_relevel(type, "True effect\n(careful responding)", "Noise\n(careless responding)", "Observed effect\n(careful + careless)"))

round_p_apa <- function(p) {
  ifelse(p < .001, "< .001", paste("= ", formatC(p, format = "f", digits = 3), sep = ""))
}

summary <- simulation$analysis_results[[plucked_iteration]] |>
  pivot_longer(cols = everything(),
               names_to = c("metric", "type"),
               values_to = "values",
               names_sep = "_") |>
  pivot_wider(names_from = "metric",
              values_from = "values") |>
  mutate(r = round_half_up(r, digits = 2),
         p = round_p_apa(p)) |>
  mutate(type = case_when(type == "careful" ~ "True effect\n(careful responding)",
                          type == "careless" ~ "Noise\n(careless responding)",
                          type == "all" ~ "Observed effect\n(careful + careless)"),
         type = fct_relevel(type, "True effect\n(careful responding)", "Noise\n(careless responding)", "Observed effect\n(careful + careless)"))

ggplot(dat_all, aes(x, y)) +
  geom_point(alpha = 0.2, color = "grey50") +
  geom_smooth(method = "lm", color = "chartreuse4") +
  coord_cartesian(xlim = c(1, 7)) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9,.15),
        #panel.spacing = unit(1, "lines"), # Adjust spacing to help equalize facet dimensions
        aspect.ratio = 1) +
  facet_wrap(~ type) +
  geom_text(data = summary, 
            aes(x = 1.25, y = 6.5, label = paste("r =", r, ", p", p)), 
            hjust = 0, 
            vjust = 0.5, 
            size = 4, 
            color = "chartreuse4")

```

# for quarto

```{r}

library(tidyr)
library(dplyr)
library(tibble)
library(forcats)
library(purrr) 
library(ggplot2)
library(janitor)
library(faux)

generate_data <- function(n,
                          prob_careless,
                          rho_careful,
                          mu_x_careful,
                          mu_y_careful) { 

  n_careless <- floor(n * prob_careless)
  n_careful <- n - n_careless
  
  data_careful <- 
    faux::rnorm_multi(n = n_careful, 
                      mu = c(y = 1, x = 1), 
                      sd = c(1, 1), 
                      r = matrix(c(1, rho_careful, 
                                   rho_careful, 1), 
                                 ncol = 2)) |>
    mutate(type = "careful") |>
    # convert to likert
    mutate(x = faux::norm2trunc(x, min = 1, max = 7, mu = mu_x_careful, sd = 1),
           y = faux::norm2trunc(y, min = 1, max = 7, mu = mu_y_careful, sd = 1))

  data_careless <-
    data.frame(x = runif(n = n_careless, min = 1, max = 7),
               y = runif(n = n_careless, min = 1, max = 7)) |>
    mutate(type = "careless") 
  
  data <- bind_rows(data_careful,
                    data_careless) |>
    rownames_to_column(var = "id")
  
  return(data)
}

# define data analysis function ----
analyse_data <- function(data) {
  
  fit_all <- cor.test(data$y,
                      data$x, 
                      method = "pearson")
  
  dat_careful <- data |>
    filter(type == "careful")
  
  fit_careful <- cor.test(dat_careful$y,
                          dat_careful$x, 
                          method = "pearson")
  
  dat_careless <- data |>
    filter(type == "careless")
  
  fit_careless <- cor.test(dat_careless$y,
                           dat_careless$x, 
                           method = "pearson")
  
  results <- tibble(r_all = fit_all$estimate,
                    p_all = fit_all$p.value,
                    r_careful = fit_careful$estimate,
                    p_careful = fit_careful$p.value,
                    r_careless = fit_careless$estimate,
                    p_careless = fit_careless$p.value,
                    meanx_careful = mean(dat_careful$x),
                    meany_careful = mean(dat_careful$y))
  
  return(results)
}

# set seed
set.seed(42)

# define experiment parameters ----
prob_careless <- 0.15

experiment_parameters_grid <- expand_grid(
  n = 800, #c(100, 200, 300),
  prob_careless = prob_careless,
  rho_careful = 0, 
  mu_x_careful = 2,
  mu_y_careful = 2
)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n,
                                    prob_careless,
                                    rho_careful,
                                    mu_x_careful,
                                    mu_y_careful),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))


plucked_iteration <- 1

dat_all <- 
  bind_rows(simulation$generated_data[[plucked_iteration]] |>
              mutate(type = case_when(type == "careful" ~ "True effect\n(careful responding)",
                                      type == "careless" ~ "Noise\n(careless responding)")),
            simulation$generated_data[[plucked_iteration]] |>
              mutate(type = "Observed effect\n(careful + careless)")) |>
  mutate(type = fct_relevel(type, "True effect\n(careful responding)", "Noise\n(careless responding)", "Observed effect\n(careful + careless)"))

round_p_apa <- function(p) {
  ifelse(p < .001, "< .001", paste("= ", formatC(p, format = "f", digits = 3), sep = ""))
}

summary <- simulation$analysis_results[[plucked_iteration]] |>
  pivot_longer(cols = everything(),
               names_to = c("metric", "type"),
               values_to = "values",
               names_sep = "_") |>
  pivot_wider(names_from = "metric",
              values_from = "values") |>
  mutate(r = round_half_up(r, digits = 2),
         p = round_p_apa(p)) |>
  mutate(type = case_when(type == "careful" ~ "True effect\n(careful responding)",
                          type == "careless" ~ "Noise\n(careless responding)",
                          type == "all" ~ "Observed effect\n(careful + careless)"),
         type = fct_relevel(type, "True effect\n(careful responding)", "Noise\n(careless responding)", "Observed effect\n(careful + careless)"))

# p1 <- 
#   ggplot(dat_all, aes(x, y)) +
#   geom_point(alpha = 0.2, color = "grey50") +
#   geom_smooth(method = "lm", color = "chartreuse4") +
#   coord_cartesian(xlim = c(1, 7)) +
#   scale_color_viridis_d(begin = 0.4, end = 0.6) +
#   theme_linedraw() +
#   theme(legend.position = "inside", 
#         legend.position.inside = c(.9,.15),
#         #panel.spacing = unit(1, "lines"), # Adjust spacing to help equalize facet dimensions
#         aspect.ratio = 1) +
#   facet_wrap(~ type) +
#   geom_text(data = summary, 
#             aes(x = 1.25, y = 6.5, label = paste("r =", r, ", p", p)), 
#             hjust = 0, 
#             vjust = 0.5, 
#             size = 4, 
#             color = "chartreuse4")


ggplot(dat_all, aes(x, y)) +
  geom_hline(yintercept = 4, linetype = "dotted") +
  geom_vline(xintercept = 4, linetype = "dotted") +
  geom_hline(data = summary |> drop_na(), 
             aes(yintercept = meany), linetype = "dotted", color = "purple") +
  geom_vline(data = summary |> drop_na(), 
             aes(xintercept = meanx), linetype = "dotted", color = "purple") +
  geom_point(alpha = 0.4, color = "grey50") +
  geom_smooth(method = "lm", color = "chartreuse4", fullrange = TRUE) +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  geom_richtext(data = summary, 
                aes(x = 1.25, y = 6.5, label = paste("<i>r</i> =", r, ", <i>p</i>", p)), 
                hjust = 0, 
                vjust = 0.5, 
                size = 8, 
                color = "chartreuse4") +
  scale_y_continuous(breaks = c(1,2,3,4,5,6,7)) +
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7)) +
  coord_cartesian(xlim = c(1, 7), ylim = c(1, 7)) +
  xlab("Variable X") +
  ylab("Variable Y") +
  facet_wrap(~ type) +
  theme_linedraw() +
  theme(legend.position = "inside", 
        legend.position.inside = c(.9, .15),
        panel.spacing = unit(1.5, "lines"))

```

# Session info

```{r}

sessionInfo()

```


