---
title: "Multiverse plots for simulation studies"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

# Simulation

Proportion of significant results from students t test under various conditions.

```{r}

# dependencies
library(tidyr)
library(dplyr)
library(purrr) 
library(ggplot2)
library(ggtext)
library(sn)
library(knitr)
library(kableExtra)
library(janitor)

# set the seed ----
# for the pseudo random number generator to make results reproducible
set.seed(42)

# define data generating function ----
generate_data <- function(n_control,
                          n_intervention,
                          location_control, # location, akin to mean
                          location_intervention,
                          scale_control, # scale, akin to SD
                          scale_intervention,
                          skew_control, # slant/skew. When 0, produces normal/gaussian data
                          skew_intervention) {
  
  data_control <- 
    tibble(condition = "control",
           score = rsn(n = n_control, 
                       xi = location_control, # location, akin to mean
                       omega = scale_control, # scale, akin to SD
                       alpha = skew_control)) # slant/skew. When 0, produces normal/gaussian data
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rsn(n = n_intervention, 
                       xi = location_intervention, # location, akin to mean
                       omega = scale_intervention, # scale, akin to SD
                       alpha = skew_intervention)) # slant/skew. When 0, produces normal/gaussian data
  
  data <- bind_rows(data_control,
                    data_intervention) 
  
  return(data)
}


# define data analysis function ----
analyse_data <- function(data) {
  res_t_test <- t.test(formula = score ~ condition, 
                       data = data,
                       var.equal = FALSE,
                       alternative = "two.sided")
  
  res <- tibble(p = res_t_test$p.value)
  
  return(res)
}

# define experiment parameters ----
experiment_parameters_grid <- expand_grid(
  n_control = c(50, 100, 150),
  #n_intervention = 100,
  location_control = 0,
  location_intervention = c(0.0, 0.2, 0.5, 0.8), 
  scale_control = 1,
  scale_intervention = c(1, 1.5),
  skew_control = 0,
  skew_intervention = 0, 
  iteration = 1:1000
) |>
  mutate(n_intervention = n_control)

# run simulation ----
simulation <- 
  # using the experiment parameters
  experiment_parameters_grid |>
  
  # generate data using the data generating function and the parameters relevant to data generation
  mutate(generated_data = pmap(list(n_control,
                                    n_intervention,
                                    location_control,
                                    location_intervention,
                                    scale_control,
                                    scale_intervention,
                                    skew_control,
                                    skew_intervention),
                               generate_data)) |>
  
  # apply the analysis function to the generated data using the parameters relevant to analysis
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data))

```

# Summarize over iterations

## Table of results

The typical way of illustrating simulation results.

```{r}

simulation_summary <- simulation |>
  unnest(analysis_results) |>
  group_by(n_control,
           n_intervention,
           location_control,
           location_intervention,
           scale_control,
           scale_intervention,
           skew_control,
           skew_intervention) |>
  summarize(positive_rate = janitor::round_half_up(mean(p < .05), digits = 3)) |>
  ungroup()

simulation_summary |>
  kable() |>
  kable_classic(full_width = FALSE)

```

## Multiverse plot function

```{r}

multiverse_plot <- function(data, 
                            outcome = "outcome", 
                            outcome_name = "Outcome",
                            interval_lower = NULL,
                            interval_upper = NULL,
                            rank_by = "rank",
                            rank_by_outcome = FALSE,
                            relative_height_of_upper_plot = 0.70, 
                            outcome_cutoff = NULL){
  
  # dependencies
  require(dplyr)
  require(tidyr)
  require(ggplot2)
  require(ggtext)
  
  # TODO consider checking inputs with rlang::ensym(variable)

  # rename outcome variable
  data <- data |>
    rename(outcome = {{outcome}},
           rank = {{rank_by}})
  
  # if intervals aren't specified, set them to the same value as outcome (so they are invisible in the plot)
  if(!is.null(interval_lower)){
    data <- data |>
      rename(interval_lower = {{interval_lower}})
  } else {
    data$interval_lower <- data$outcome
  }
  
  if(!is.null(interval_upper)){
    data <- data |>
      rename(interval_upper = {{interval_upper}})
  } else {
    data$interval_upper <- data$outcome
  }
  
  # if rank_by_outcome, rank by the outcome. otherwise, take order that was passed.
  if(rank_by_outcome){
    data <- data |>
      arrange(outcome) |>
      mutate(rank = row_number())
  }
  
  # ensure all columns other than rank and outcome are factors
  data <- data |>
    mutate(across(.cols = -c(rank, outcome, interval_lower, interval_upper),
                  .fns = as.factor))
  
  p_estimates <- data |>
    mutate(outcome_name = outcome_name) |>
    ggplot(aes(rank, outcome)) +
    geom_linerange(aes(ymin = interval_lower, ymax = interval_upper)) +
    geom_point(shape = "circle", size = 2) +
    facet_grid(outcome_name ~ ., space = "free_y", scales = "free_y", switch = "y") +
    scale_y_continuous(breaks = scales::pretty_breaks()) + 
    scale_x_continuous(NULL, expand = c(.02, .02)) +
    ylab("Simulation results") +
    #theme_classic() +
    theme_minimal() +
    theme(legend.position = "none",
          axis.line.x = element_blank(),
          strip.placement = "outside",
          strip.background = element_rect(fill = NA, colour = NA),
          panel.spacing.x = unit(0.15, "cm"),
          strip.text.y = element_markdown(angle = 180, face = "bold", size = 7),
          axis.text.y = element_text(angle = 0, 
                                     hjust = 0.5, 
                                     size = 6),
          axis.title.y = element_text(size = 9, face = "bold"),
          panel.spacing = unit(0.25, "lines"),
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
  
  if(!is.null(outcome_cutoff)){
    p_estimates <- p_estimates + geom_hline(yintercept = outcome_cutoff, linetype = "dotted")
  }
  
  p_specs <- data |> 
    tidyr::pivot_longer(cols = c(-"rank", -"outcome", -"interval_lower", -"interval_upper")) |> 
    arrange(rank) |>
    ggplot(aes(x = rank, y = factor(value), color = name)) + 
    geom_point(size = 2, shape = "square") +
    facet_grid(name ~ ., space = "free_y", scales = "free_y", switch = "y") +
    guides(color = "none") +
    scale_x_continuous(NULL, expand = c(.02, .02)) +
    ylab("Simulation conditions") +
    #theme_classic() +
    theme_minimal() +
    theme(strip.placement = "outside",
          strip.background = element_rect(fill = NA, colour = NA),
          panel.spacing.x = unit(0.15, "cm"),
          strip.text.y = element_markdown(angle = 180, face = "bold", size = 7),
          axis.text.y = element_text(angle = 0, hjust = 1, size = 6),
          axis.title.y = element_text(size = 9, face = "bold"),
          panel.spacing = unit(0.25, "lines"),
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) +
    scale_color_brewer(palette = "Dark2")

  p_multiverse <- cowplot::plot_grid(p_estimates, 
                                     p_specs, 
                                     axis = "bltr", 
                                     align = "v", 
                                     ncol = 1, 
                                     rel_heights = c(relative_height_of_upper_plot, 1))
  
  return(p_multiverse)
}

```

### Example 1

Left-right order of conditions determined by their permutation.

```{r}

data_summary_ordered <- simulation_summary |>
  # the outcome variable must be called `outcome`
  # eg the estimand of your simulation, whether the false positive rate, estimated effect size, etc.
  # replace `positive_rate` with
  mutate(outcome = positive_rate) |>
  
  # order the rows within the columns in the way you want them to appear left to right
  arrange(location_intervention, n_intervention, scale_intervention) |>
  mutate(rank = row_number()) |>
  
  # # note that you might instead want to rank by the outcome variable rather than the conditions
  # arrange(outcome) |>
  # mutate(rank = row_number())
  
  # select only the columns that you had multiple values for in the expand.grid() call in the simulation as well as `outcome` and `rank`
  # rename them if necessary to plot nicely
  dplyr::select(`N per group` = n_control, 
                #n_intervention, 
                #location_control, 
                `Population mean` = location_intervention, 
                #scale_control, 
                Variances = scale_intervention, 
                #skew_control, 
                #skew_intervention, 
                outcome,
                rank) |>
  # if it would look better for plotting, you can recategorize some values as named factor levels
  mutate(Variances = case_when(Variances == 1 ~ "Equal",
                               Variances == 1.5 ~ "Unequal"))

p_multiverse <- 
  multiverse_plot(data_summary_ordered, 
                  outcome_name = "Proportion p < .05")

p_multiverse

# # save to disk
# ggsave(filename = "p_multiverse.pdf",
#        plot = p_multiverse,
#        device = "pdf",
#        width = 6,
#        height = 5)

```

### Example 2

Left-right order of conditions still determined by their permutation. Also add a cutoff line, eg for "acceptable power" (>=.80).

```{r}

multiverse_plot(data_summary_ordered, 
                outcome_name = "Proportion p < .05",
                outcome_cutoff = 0.80)

```

### Example 3

Left-right order of conditions still determined by their permutation. Add intervals on the estimates.

```{r}

data_summary_ordered |>
  # create some dummy intervals to show they can be plotted too. doesn't make sense for p values, its just illustrative
  mutate(interval_lower = outcome - 0.3, 
         interval_upper = outcome + 0.3) |>
  # plot
  multiverse_plot(outcome_name = "Proportion p < .05",
                  relative_height_of_upper_plot = 1, # increase size of upper plot 
                  interval_lower = "interval_lower", # specify the interval columns
                  interval_upper = "interval_upper")

```

### Example 4

Left-right order of conditions determined by the ranked outcome variable. 

```{r}

data_summary_ordered <- simulation_summary |>
  # the outcome variable must be called `outcome`
  # eg the estimand of your simulation, whether the false positive rate, estimated effect size, etc.
  # replace `positive_rate` with
  mutate(outcome = positive_rate) |>
  
  # # note that you might instead want to rank by the outcome variable rather than the conditions
  # arrange(outcome) |>
  # mutate(rank = row_number())
  
  # select only the columns that you had multiple values for in the expand.grid() call in the simulation as well as `outcome` and `rank`
  # rename them if necessary to plot nicely
  dplyr::select(
    `N per group` = n_control, 
    #n_intervention, 
    #location_control, 
    Variances = scale_intervention, 
    #scale_control, 
    `Effect size` = location_intervention, 
    #skew_control, 
    #skew_intervention, 
    outcome
  ) |>
  # if it would look better for plotting, you can recategorize some values as named factor levels
  mutate(Variances = case_when(Variances == 1 ~ "Equal",
                               Variances == 1.5 ~ "Unequal")) |>
  
  # order the rows within the columns in the way you want them to appear left to right
  arrange(
    `Effect size`, 
    `N per group`, 
    Variances,
  ) |>
  mutate(rank = row_number())

p_current <- 
  multiverse_plot(data_summary_ordered, 
                  rank_by_outcome = TRUE,
                  outcome_name = "Proportion p < .05")

p_current

# ggsave(filename = "p_3.pdf",
#        plot = p_current,
#        device = "pdf",
#        width = 6,
#        height = 5)

```





