

# simple demo of collider

```{r}

library(lavaan)
library(semPlot)
library(tidyverse)

analyse <- function(model, data){
  # fit regressions. NB model must include a Y ~ X term.
  fit <- sem(model = model, data = data) 
  
  # extract beta estimate for Y~X
  beta <- parameterEstimates(fit) |>
    mutate(effect = paste(lhs, rhs, sep = "~")) |>
    filter(effect == "Y~X") |>
    dplyr::select(y_x_beta = est)
  
  fit_measures <- data.frame(estimate = fitmeasures(fit, fit.measures = c("rmsea", "tli", "cfi", "srmr"))) |>
    rownames_to_column(var = "fit_metric") |>
    pivot_wider(names_from = fit_metric, 
                values_from = estimate)
  
  return(bind_cols(beta, fit_measures))
}

```

# Collider analyzed as if its multiple regression

true Y~X is zero, but analysis shows a non-zero Y~X effect!

despite fitting a totally different analytic model to the data generating process, we find good fit!

```{r}

# data generating process
mod <- "M ~ 0.5*X + 0.5*Y; Y ~ 0.0*X"

# simulate data
dat <- lavaan::simulateData(model = mod, sample.nobs = 10000)

# plot model configuration
layout_matrix <- matrix(c( 0,  1,
                           1,  0,
                          -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

dat %>%
  sem(model = mod, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Collider", adj = 0)

# analyze using a multiple regression model
analyse(model = "Y ~ X + M", data = dat)

```

# Collider analyzed as if its multiple regression (ii)

true Y~X is positive (0.5), but analysis shows an attenuated Y~X effect!

despite fitting a totally different analytic model to the data generating process, we find good fit!

```{r}

# data generating process
mod <- "M ~ 0.5*X + 0.5*Y; Y ~ 0.5*X"

# simulate data
dat <- lavaan::simulateData(model = mod, sample.nobs = 10000)

# plot configural model
layout_matrix <- matrix(c( 0,  1,
                           1,  0,
                          -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

dat %>%
  sem(model = mod, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Collider", adj = 0)

# analyze using a multiple regression model
analyse(model = "Y ~ X + M", data = dat)

```

# Collider analyzed as if its mediation

when true Y~X is positive (o.5), results can 'support' partial mediation! ie a positive direct effect reduces when you control for the mediator

```{r}

# data generating process
mod <- "M ~ 0.5*X + 0.5*Y; Y ~ 0.5*X"

# simulate data
dat <- lavaan::simulateData(model = mod, sample.nobs = 10000)

dat %>%
  sem(model = mod, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Collider", adj = 0)

# direct effect `c`
analyse(model = "Y ~ X", data = dat)
# direct effect `c'` after controlling for mediator
analyse(model = "Y ~ X + M; M ~ X", data = dat)

```

- bring up Baron & Kenny's article, how often its cited, and the influence it has on the literature in the last few decades
- bring up how this method seems to align with the goals of psychology wanting to know about mental processes.
- the point is that "mediation tests" do not test mediation, they assume mediation.



