---
title: "Regression assumes causality"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_download: true
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# Citation & License

Citation: 

Ian Hussey (2024) Improving your statistical inferences through simulation studies in R. https://github.com/ianhussey/simulation-course

License: 

[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/deed.en)

```{r setup, include=FALSE}

knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen=999)

```

# Dependencies

```{r}

library(tidyr)
library(dplyr)
library(purrr) 
library(parameters)
library(lavaan)
library(semPlot)
library(knitr)
library(kableExtra)

```

# Functions

```{r}

generate_data <- function(n, population_model) {
  
  data <- lavaan::simulateData(model = population_model, sample.nobs = n) 
  
  return(data)
}

analyse <- function(data, model) {
  
  # specify and fit model
  fit <- sem(model = model, data = data)
  
  # extract regression beta estimates 
  results <- model_parameters(fit, standardize = TRUE) |>
    filter(To == "Y" & From == "X") |>  # this corresponds to the Y ~ X effect
    select(beta = Coefficient,
           ci_lower = CI_low,
           ci_upper = CI_high,
           p) 
  
  return(results)
}

plots_causal <- function(model){
  generate_data(n = 300, population_model = model) %>%
    sem(model = model, data = .) |>
    semPaths(whatLabels = "diagram", 
             layout = layout_matrix, 
             residuals = FALSE,
             edge.label.cex = 1.2, 
             sizeMan = 10)
}

```

# Does regression assume causality?

## Plots

Simple regression: X causes Y

```{r}

# simple regression
layout_matrix <- matrix(c( 1,  0,
                           -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

plots_causal("Y ~ 0.5*X")

```

Simple regression: Y causes X

```{r}

# simple regression
layout_matrix <- matrix(c(-1,  0,
                           1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

plots_causal("X ~ 0.5*Y")

```

## Run simulation

```{r}

experiment_parameters_grid <- expand_grid(
  n = 200,
  population_model = c("Y ~ 0.5*X",
                       "X ~ 0.5*Y"),
  analyse_model = "Y ~ X",
  iteration = 1:1000
) 

set.seed(42)

simulation <- 
  # using the experiment parameters...
  experiment_parameters_grid |>
  
  # ...generate data... 
  mutate(generated_data = pmap(list(n = n,
                                    population_model = population_model),
                               generate_data)) |>
  # ...analyze data 
  mutate(results = pmap(list(data = generated_data,
                             model = analyse_model),
                        analyse))

```

## Summarize results

```{r}

simulation_summary <- simulation |>
  unnest(results) |>
  group_by(n,
           population_model,
           analyse_model) |>
  summarize(mean_beta = mean(beta),
            proportion_signficiant = mean(p < .05))

simulation_summary |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Collider

## Plots

Covariate is a confounder:

```{r}

layout_matrix <- matrix(c( 1,  0,
                           -1,  0,
                           0,  1), 
                        ncol = 2, 
                        byrow = TRUE)

plots_causal("Y ~ 0.0*X + 0.5*C; X ~ 0.5*C")

```

Confounder is a collider:

```{r}

layout_matrix <- matrix(c( 0,  1,
                           1,  0,
                           -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

plots_causal("C ~ 0.5*X + 0.5*Y; Y ~ 0.0*X")

```

## Run simulation

```{r}

experiment_parameters_grid <- expand_grid(
  n = 200,
  population_model = c("Y ~ 0.5*X + 0.0*C",
                       "C ~ 0.5*X + 0.5*Y; Y ~~ 0.0*X"),
  analyse_model = "Y ~ X + C",
  iteration = 1:1000
)

set.seed(42)

simulation <- 
  # using the experiment parameters...
  experiment_parameters_grid |>
  
  # ...generate data...
  mutate(generated_data = pmap(list(n = n,
                                    population_model = population_model),
                               generate_data)) |>
  # ...analyze data
  mutate(results = pmap(list(data = generated_data,
                             model = analyse_model),
                        analyse))

```

## Summarize results

```{r}

simulation_summary <- simulation |>
  unnest(results) |>
  group_by(n,
           population_model,
           analyse_model) |>
  summarize(mean_beta = mean(beta),
            proportion_signficiant = mean(p < .05))

simulation_summary |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Session info

```{r}

sessionInfo()

```


