---
title: "Practice writing R functions"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_download: true
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# Overview

Two of the key steps in a simulation study (generate data and analyze data) require us to know how to write functions. This R Markdown lesson practices this.

```{r, include=FALSE}

# set default chunk options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

# dependencies
library(dplyr)
library(tibble)
library(forcats)
library(report)

```

# Primer on functions

Most code we use are functions, e.g., `mean()`, `setwd()` and `library()`.

These functions were written by others, but we can write our own. 

"It's functions all the down": you will use existing functions to write new ones. For example:

```{r}

values <- c(4, 2, 6, 2, NA, 4, 3, 1, NA, 7, 5)

mean(values) # returns NA 
mean(values, na.rm = TRUE) # returns the mean after dropping NA

# tired of writing 'na.rm = TRUE' repeatedly? write your own function to do it automatically
mean_na_rm <- function(x){
  mean(x, na.rm = TRUE)
}

mean_na_rm(values) # returns the mean after dropping NA

```

What if we usually want to `round()` to two decimal places, and we're tired of writing `digits = 2` every time?

```{r}

mean_of_values <- mean_na_rm(values)

round(mean_of_values, digits = 2)

# write a function to always round to two decimal places
round_2 <- function(x){
  round(x, digits = 2)
}

round_2(mean_of_values)

```

## Mini lesson: round() probably doesn't do what you think it does

round() uses "banker's rounding" rather than the round-half-up method we're used to

```{r}

round(0.5)
round(1.5)
round(2.5)
round(3.5)
round(4.5)
round(5.5)

```


```{r}

library(janitor)

round_half_up(0.5)
round_half_up(1.5)
round_half_up(2.5)
round_half_up(3.5)
round_half_up(4.5)
round_half_up(5.5)

```

## General structure of a function

Functions (usually) have 'inputs', they have code that they run ('do stuff'), and they (almost always) return 'outputs'. The often specify their requirements and include checks that their inputs are correctly formatted.

Note that this is pseudo-code only: chunk is set not to run (`eval=FALSE`).

```{r eval=FALSE}

# define function
function_name <- function(argument_1, # first argument is often the data, if the function takes a data frame as an argument
                          argument_2 = "default", # arguments can have defaults
                          argument_3) {
  # required packages
  require(dplyr)
  
  # checks
  # well written functions contain checks. 
  # e.g., if the function assumes that argument_1 is a data frame, check that this is the case.
  # note that it is more useful to write the function first and add checks later.
  if(!is.data.frame(argument_1)){
    stop("argument_1 must be a data frame")
  }
  
  # code that does things
  object_to_be_returned <- input_data_frame |>
    # do things
    mutate(value = value + 1)
  
  # object to be returned
  return(object_to_be_returned)
}

```

## Example function: t-test p value

```{r}

# use this data to build and test your function
data_simulated_intervention <- 
  tibble(condition = "intervention", 
         score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated_control <- 
  tibble(condition = "control", 
         score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated <- 
  bind_rows(data_simulated_intervention,
            data_simulated_control)

```

```{r}

# define function
t_test_p_value <- function(data) {

  res <- t.test(formula = score ~ condition, 
                data = data)
  
  return(res$p.value)
}

# call function
t_test_p_value(data_simulated)

```

## General things to remember when writing functions

- If you can't immediately write the code, write pseudo-code first! 
- Build the 'do stuff' part outside of a function first!
- Wrap the 'do stuff' with input and output after you have 'do stuff' working. Why: so you don't have to fight variable scoping.
- The function must be present in your environment to be usable, and must be called to be used
- Check that your function actually works as you expect, not just that it runs. Give it lots of different input values that should and should not work, and check you get the correct outputs.
- Don't try to abstract more than you need. One function should do one thing. Elaborate the function only as needed.

# Practice writing functions

## Calculate mean dropping NAs

```{r}



```

## Calculate SD dropping NAs

```{r}



```

## Calculate mean for each condition dropping NAs

```{r}



```

## Calculate mean and SD for each condition dropping NAs

```{r}



```

## Calculate mean and SD for each condition dropping NAs, then use them to calculate Cohen's d

```{r}



```

## p value from correlation test

Goal: Fit a correlation test using cor.test() and extract the p value and correlation

```{r}

# use this data to build and test your function
data_simulated <- tibble(Y = rnorm(n = 100, mean = 0, sd = 1),
                         X = rnorm(n = 100, mean = 0, sd = 1))


cor_extraction <- function(data){
  
  # do stuff
  
  return(res)
}

cor_extraction(data_simulated)

```

## Pearson's r from correlation test

Goal: Fit a correlation test using cor.test() and extract the p value and correlation

```{r}

# use this data to build and test your function
data_simulated <- tibble(Y = rnorm(n = 100, mean = 0, sd = 1),
                         X = rnorm(n = 100, mean = 0, sd = 1))


cor_extraction <- function(data){
  
  # do stuff
  
  return(res)
}

cor_extraction(data_simulated)

```

## Pearson's r and its p value from correlation test

Goal: Fit a correlation test using cor.test() and extract the p value and correlation

```{r}

# use this data to build and test your function
data_simulated <- tibble(Y = rnorm(n = 100, mean = 0, sd = 1),
                         X = rnorm(n = 100, mean = 0, sd = 1))


cor_extraction <- function(data){
  
  # do stuff
  
  return(res)
}

cor_extraction(data_simulated)

```

## Cohen's d and its confidence intervals 

Goal: Calculate cohen's d using effsize::cohen.d() and return the estimate and its p values as a data frame.

```{r}

# use this data to build and test your function
data_simulated_intervention <- tibble(condition = "intervention", 
                                      score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated_control <- tibble(condition = "control", 
                                 score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated <- 
  bind_rows(data_simulated_intervention,
            data_simulated_control) |>
  # control's factor levels must be ordered so that intervention is the first level and control is the second
  # this ensures that positive cohen's d values refer to intervention > control and not the other way around.
  mutate(condition = fct_relevel(condition, "intervention", "control"))



extract_d <- function(data){
  
  # do stuff
  
  return(res)
}

extract_d(data_simulated)

```

## t-test's p value, Cohen's d and its confidence intervals

```{r}

# use this data to build and test your function
data_simulated_intervention <- tibble(condition = "intervention", 
                                      score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated_control <- tibble(condition = "control", 
                                 score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated <- 
  bind_rows(data_simulated_intervention,
            data_simulated_control) |>
  # control's factor levels must be ordered so that intervention is the first level and control is the second
  # this ensures that positive cohen's d values refer to intervention > control and not the other way around.
  mutate(condition = fct_relevel(condition, "intervention", "control"))



analyze_data <- function(data) {
  
  # do stuff
  
  return(res)
}

```

## Generate data for a between groups design

Goal: rather than writing a data analysis function, this time write a data generation function. In the previous chunks we've used this code to generate a single data set with intervention and control conditions and simulated normally distributed data. Rewrite this as a function so that we can generate such a data set with one line of code using the new function `generate_data()`. Unlike your previous functions, this one has no inputs, i.e., you can write `function() <- `.

```{r}

# use this *code* to build your data generation function
data_simulated_intervention <- tibble(condition = "intervention", 
                                      score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated_control <- tibble(condition = "control", 
                                 score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated <- 
  bind_rows(data_simulated_intervention,
            data_simulated_control) |>
  # control's factor levels must be ordered so that intervention is the first level and control is the second
  # this ensures that positive cohen's d values refer to intervention > control and not the other way around.
  mutate(condition = fct_relevel(condition, "intervention", "control"))




generate_data <- function(){
  
  # do stuff
  
  return(data_simulated)
}

generate_data() |>
  analyze_data()

```

## Create a more flexible data generation function 

Goal: What if we want to generate data with a different number of participants per condition? We can make the value of `n` a variable rather than a hard-coded value and add this to functions inputs/arguments.

```{r}

generate_data <- function(n_per_group){
  
  # do stuff
  
  return(data_simulated)
}

bind_rows(
  generate_data(n_per_group = 50) |>
    analyze_data(),
  generate_data(n_per_group = 100) |>
    analyze_data(),
  generate_data(n_per_group = 150) |>
    analyze_data(),
  generate_data(n_per_group = 200) |>
    analyze_data(),
  generate_data(n_per_group = 250) |>
    analyze_data(),
) |>
  mutate(ci_width = cohens_d_ci_upper - cohens_d_ci_lower) |>
  select(ci_width)

```

## Create an even more flexible data generation function 

Goal: Rewrite the data generation function to make it even more flexible: make all Ns, Ms, and SDs variables that are set by the inputs.

```{r}

generate_data <- function(n_intervention, n_control = NA, m_intervention = 0, m_control = 0, sd_intervention = 1, sd_control = 1){
  
  # do stuff
  
  return(data_simulated)
}

generate_data(n_intervention = 25, 
              n_control = 25, 
              m_intervention = .5)

```

# Notes

Although we have practiced writing custom functions to extract statistical results / model parameters, it is worth knowing that the easystats family of packages includes [{parameters}](https://easystats.github.io/parameters/) package, which does a very good job of extracting model parameters from a very wide range of models including base R functions, lavaan, psych, and other packages. If you want to extract values from a model, consider using {parameters} to do a lot of the work for you when writing your function.

Seperately, the [{report}](https://easystats.github.io/report/) package will fully report the results of many common analyses for you. e.g.:

```{r}

data_simulated_intervention <- 
  tibble(condition = "intervention", 
         score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated_control <- 
  tibble(condition = "control", 
         score = rnorm(n = 50, mean = 0, sd = 1))

data_simulated <- 
  bind_rows(data_simulated_intervention,
            data_simulated_control)

t.test(score ~ condition, data = data_simulated) |>
  report::report()

```

This lesson does not cover documenting your functions well, organizing them into an R package to make them easy to load and include help menus, or writing unit tests them. These are all very worth doing. Look into the {roxygen} package.

# Session info

```{r}

sessionInfo()

```


