---
title: "The differences between significant and non-significant is not itself significant"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_download: true
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# Overview of tutorial

XX

# Citation & License

Citation: 

Ian Hussey (2024) Improving your statistical inferences through simulation studies in R. https://github.com/ianhussey/simulation-course

License: 

[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/deed.en)

```{r, include=FALSE}

# set default chunk options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# Dependencies

```{r}

library(tidyr)
library(dplyr)
library(purrr) 
library(stringr)
library(forcats)
library(ggplot2)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)
library(janitor)
library(effsize)
library(metafor)

format_p_apa <- function(p, threshold = 0.001) {
  ifelse(
    p < threshold,
    paste0("p < ", sub("^0\\.", ".", format(threshold, nsmall = 3))),
    paste0("p = ", sub("^0\\.", ".", format(round(p, 3), nsmall = 3)))
  )
}

```

# Simulation functions

```{r}

# functions for simulation
generate_data <- function(n_per_condition,
                          population_cohens_d) {
  
  data_control <- 
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = 0, sd = 1))
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = population_cohens_d, sd = 1))
  
  data_combined <- bind_rows(data_control,
                             data_intervention) |>
    mutate(condition = fct_relevel(condition, "intervention", "control"))
  
  return(data_combined)
}

analyze <- function(data, study) {

  res_t_test <- t.test(formula = score ~ condition, 
                       data = data,
                       var.equal = TRUE,
                       alternative = "two.sided")
  
  res_cohens_d <- effsize::cohen.d(formula = score ~ condition,
                                   data = data,
                                   pooled = TRUE)
  
  res <- tibble(study = study,
                p = res_t_test$p.value, 
                cohens_d = res_cohens_d$estimate,
                cohens_d_ci_lower = res_cohens_d$conf.int[1],
                cohens_d_ci_upper = res_cohens_d$conf.int[2],
                cohens_d_se = sqrt(res_cohens_d$var))

  return(res)
}

meta_analyze <- function(results_study_1, results_study_2) {
  
  dat <- bind_rows(results_study_1,
                   results_study_2)
  
  fit <- rma(data = dat, 
             yi = cohens_d, 
             sei = cohens_d_se, 
             mods = ~ study, 
             method = "FE")

  res <- tibble(cohens_d = fit$b[2],
                cohens_d_ci_lower = fit$ci.lb[2],
                cohens_d_ci_upper = fit$ci.ub[2],
                #df = fit$QMdf[1],
                #Q = fit$QM,
                p = fit$QMp) # p of moderation
    #mutate(formatted = paste0("Q(", df, ") = ", round_half_up(Q, 3), ", ", format_p_apa(p)))

  return(res)
}

```

# Different population effect, large N

```{r fig.height=5, fig.width=14}

# set seed
set.seed(42)

# simulation parameters
experiment_parameters <- expand_grid(
  study_1_n_per_condition = 300,
  study_1_population_cohens_d = 0.5,
  study_2_n_per_condition = 300,
  study_2_population_cohens_d = 0,
  iteration = 1:1000
) 

# run simulation
simulation <- experiment_parameters |>
  mutate(generated_data_study_1 = pmap(list(n_per_condition = study_1_n_per_condition, 
                                            population_cohens_d = study_1_population_cohens_d),
                                       generate_data)) |>
  mutate(generated_data_study_2 = pmap(list(n_per_condition = study_2_n_per_condition, 
                                            population_cohens_d = study_2_population_cohens_d),
                                       generate_data)) |>
  mutate(results_study1 = pmap(list(data = generated_data_study_1),
                                analyze,
                                study = "study 1")) |>
  mutate(results_study2 = pmap(list(data = generated_data_study_2),
                                analyze,
                                study = "study 2")) |>
  mutate(results_meta = pmap(list(results_study_1 = results_study1,
                                  results_study_2 = results_study2),
                             meta_analyze)) |>
  # wrangle results
  unnest(results_study1, names_sep = "_") |>
  unnest(results_study2, names_sep = "_") |>
  unnest(results_meta, names_sep = "_") |>
  select(-generated_data_study_1, -generated_data_study_2) |>
  # this is a complex pivot, don't worry if you don't immediately understand it
  pivot_longer(
    cols = c(
      results_study1_p,
      results_study1_cohens_d,
      results_study1_cohens_d_ci_lower,
      results_study1_cohens_d_ci_upper,
      results_study1_cohens_d_se,
      results_study2_p,
      results_study2_cohens_d,
      results_study2_cohens_d_se,
      results_study2_cohens_d_ci_lower,
      results_study2_cohens_d_ci_upper,
      results_meta_p,
      results_meta_cohens_d,
      # results_meta_cohens_d_se, # not created
      results_meta_cohens_d_ci_lower,
      results_meta_cohens_d_ci_upper
    ),
    names_to = c("source", "metric"),
    names_pattern = "results_(study\\d+|meta)_(.+)",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = metric,
    values_from = value
  )

simulation_summary <- simulation |>
  group_by(study_1_n_per_condition,
           study_1_population_cohens_d,
           study_2_n_per_condition,
           study_2_population_cohens_d,
           source) |>
  summarize(proportion_significant = mean(p < .05),
            mean_cohens_d = mean(cohens_d),
            mean_cohens_d_ci_lower = mean(cohens_d_ci_lower),
            mean_cohens_d_ci_upper = mean(cohens_d_ci_upper),
            .groups = "drop") |>
  mutate(source = str_replace(source, "meta", "difference")) |>
  mutate(source = fct_relevel(source, "study1", "study2", "difference"))


p_effect_size <- ggplot(simulation_summary, aes(source, mean_cohens_d)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_linerange(aes(ymin = mean_cohens_d_ci_lower, ymax = mean_cohens_d_ci_upper)) +
  geom_point() +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     name = "Mean Cohen's d") +
  xlab("") +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_proportion_significant <- ggplot(simulation_summary, aes(source, proportion_significant)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_pretty(n = 10), limits = c(0, 1), name = "Proportion of significant p values") +
  xlab("") +
  # geom_text(aes(x = "study1", y = 1.1, label = paste("Population\nCohen's d =", study_1_population_cohens_d))) +
  # geom_text(aes(x = "study2", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d))) +
  # geom_text(aes(x = "difference", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d - study_1_population_cohens_d))) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_effect_size + p_proportion_significant + plot_layout(ncol = 2)

```

# Different population effect, small N

```{r fig.height=5, fig.width=14}

# set seed
set.seed(42)

# simulation parameters
experiment_parameters <- expand_grid(
  study_1_n_per_condition = 40,
  study_1_population_cohens_d = 0.5,
  study_2_n_per_condition = 40,
  study_2_population_cohens_d = 0,
  iteration = 1:1000
) 

# run simulation
simulation <- experiment_parameters |>
  mutate(generated_data_study_1 = pmap(list(n_per_condition = study_1_n_per_condition, 
                                            population_cohens_d = study_1_population_cohens_d),
                                       generate_data)) |>
  mutate(generated_data_study_2 = pmap(list(n_per_condition = study_2_n_per_condition, 
                                            population_cohens_d = study_2_population_cohens_d),
                                       generate_data)) |>
  mutate(results_study1 = pmap(list(data = generated_data_study_1),
                                analyze,
                                study = "study 1")) |>
  mutate(results_study2 = pmap(list(data = generated_data_study_2),
                                analyze,
                                study = "study 2")) |>
  mutate(results_meta = pmap(list(results_study_1 = results_study1,
                                  results_study_2 = results_study2),
                             meta_analyze)) |>
  # wrangle results
  unnest(results_study1, names_sep = "_") |>
  unnest(results_study2, names_sep = "_") |>
  unnest(results_meta, names_sep = "_") |>
  select(-generated_data_study_1, -generated_data_study_2) |>
  # this is a complex pivot, don't worry if you don't immediately understand it
  pivot_longer(
    cols = c(
      results_study1_p,
      results_study1_cohens_d,
      results_study1_cohens_d_ci_lower,
      results_study1_cohens_d_ci_upper,
      results_study1_cohens_d_se,
      results_study2_p,
      results_study2_cohens_d,
      results_study2_cohens_d_se,
      results_study2_cohens_d_ci_lower,
      results_study2_cohens_d_ci_upper,
      results_meta_p,
      results_meta_cohens_d,
      # results_meta_cohens_d_se, # not created
      results_meta_cohens_d_ci_lower,
      results_meta_cohens_d_ci_upper
    ),
    names_to = c("source", "metric"),
    names_pattern = "results_(study\\d+|meta)_(.+)",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = metric,
    values_from = value
  )

simulation_summary <- simulation |>
  group_by(study_1_n_per_condition,
           study_1_population_cohens_d,
           study_2_n_per_condition,
           study_2_population_cohens_d,
           source) |>
  summarize(proportion_significant = mean(p < .05),
            mean_cohens_d = mean(cohens_d),
            mean_cohens_d_ci_lower = mean(cohens_d_ci_lower),
            mean_cohens_d_ci_upper = mean(cohens_d_ci_upper),
            .groups = "drop") |>
  mutate(source = str_replace(source, "meta", "difference")) |>
  mutate(source = fct_relevel(source, "study1", "study2", "difference"))


p_effect_size <- ggplot(simulation_summary, aes(source, mean_cohens_d)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_linerange(aes(ymin = mean_cohens_d_ci_lower, ymax = mean_cohens_d_ci_upper)) +
  geom_point() +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     name = "Mean Cohen's d") +
  xlab("") +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_proportion_significant <- ggplot(simulation_summary, aes(source, proportion_significant)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_pretty(n = 10), limits = c(0, 1), name = "Proportion of significant p values") +
  xlab("") +
  # geom_text(aes(x = "study1", y = 1.1, label = paste("Population\nCohen's d =", study_1_population_cohens_d))) +
  # geom_text(aes(x = "study2", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d))) +
  # geom_text(aes(x = "difference", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d - study_1_population_cohens_d))) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_effect_size + p_proportion_significant + plot_layout(ncol = 2)

```

# Different population effect, small original study and large "replication"

NB not really a replication as the second study has a different effect size, but its a very rough simulation assuming p hacking took place in the original study and boosted the effect size

[needs work to be clearer]

```{r fig.height=10, fig.width=7}

# set seed
set.seed(42)

# simulation parameters
experiment_parameters <- expand_grid(
  study_1_n_per_condition = 40,
  study_1_population_cohens_d = 0.5,
  study_2_n_per_condition = 200,
  study_2_population_cohens_d = 0,
  iteration = 1:1000
) 

# run simulation
simulation <- experiment_parameters |>
  mutate(generated_data_study_1 = pmap(list(n_per_condition = study_1_n_per_condition, 
                                            population_cohens_d = study_1_population_cohens_d),
                                       generate_data)) |>
  mutate(generated_data_study_2 = pmap(list(n_per_condition = study_2_n_per_condition, 
                                            population_cohens_d = study_2_population_cohens_d),
                                       generate_data)) |>
  mutate(results_study1 = pmap(list(data = generated_data_study_1),
                                analyze,
                                study = "study 1")) |>
  mutate(results_study2 = pmap(list(data = generated_data_study_2),
                                analyze,
                                study = "study 2")) |>
  mutate(results_meta = pmap(list(results_study_1 = results_study1,
                                  results_study_2 = results_study2),
                             meta_analyze)) |>
  # wrangle results
  unnest(results_study1, names_sep = "_") |>
  unnest(results_study2, names_sep = "_") |>
  unnest(results_meta, names_sep = "_") |>
  select(-generated_data_study_1, -generated_data_study_2) |>
  # this is a complex pivot, don't worry if you don't immediately understand it
  pivot_longer(
    cols = c(
      results_study1_p,
      results_study1_cohens_d,
      results_study1_cohens_d_ci_lower,
      results_study1_cohens_d_ci_upper,
      results_study1_cohens_d_se,
      results_study2_p,
      results_study2_cohens_d,
      results_study2_cohens_d_se,
      results_study2_cohens_d_ci_lower,
      results_study2_cohens_d_ci_upper,
      results_meta_p,
      results_meta_cohens_d,
      # results_meta_cohens_d_se, # not created
      results_meta_cohens_d_ci_lower,
      results_meta_cohens_d_ci_upper
    ),
    names_to = c("source", "metric"),
    names_pattern = "results_(study\\d+|meta)_(.+)",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = metric,
    values_from = value
  )

simulation_summary <- simulation |>
  group_by(study_1_n_per_condition,
           study_1_population_cohens_d,
           study_2_n_per_condition,
           study_2_population_cohens_d,
           source) |>
  summarize(proportion_significant = mean(p < .05),
            mean_cohens_d = mean(cohens_d),
            mean_cohens_d_ci_lower = mean(cohens_d_ci_lower),
            mean_cohens_d_ci_upper = mean(cohens_d_ci_upper),
            .groups = "drop") |>
  mutate(source = str_replace(source, "meta", "difference")) |>
  mutate(source = fct_relevel(source, "study1", "study2", "difference"))


p_effect_size <- ggplot(simulation_summary, aes(source, mean_cohens_d)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_linerange(aes(ymin = mean_cohens_d_ci_lower, ymax = mean_cohens_d_ci_upper)) +
  geom_point() +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     name = "Mean Cohen's d") +
  xlab("") +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_proportion_significant <- ggplot(simulation_summary, aes(source, proportion_significant)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_pretty(n = 10), limits = c(0, 1), name = "Proportion of significant p values") +
  xlab("") +
  # geom_text(aes(x = "study1", y = 1.1, label = paste("Population\nCohen's d =", study_1_population_cohens_d))) +
  # geom_text(aes(x = "study2", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d))) +
  # geom_text(aes(x = "difference", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d - study_1_population_cohens_d))) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_effect_size + p_proportion_significant + plot_layout(ncol = 1)

```



# Session info

```{r}

sessionInfo()

```


